---
title: "Dispersal exploration"
subtitle: "Population fluctuations under different dispersal scenarios"
author: "J. N. Thiem"
bibliography: fragmentation.bib
csl: apa-eu.csl
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    respect-user-color-scheme: true
    theme: 
      light: flatly
      dark: darkly
    highlight-style: 
      light: github-light
      dark: github-dark
execute:
  echo: true
  warning: false
  message: false
---

## Background

The simulations used for Gelber et al. [@gelber_geometric_2025] all used an exponential dispersal kernel, meaning that the probability of dispersal decreases exponentially with distance. This is a common assumption in population ecology and should be a fair representation of dispersal for many sessile organisms (e.g. seeds dispersed by wind). Mean dispersal distance was varied between different simulations, to see how it might affect geometric and demographic effects of fragmentation. At the same time, the study also tried to determine how different edge effects (from strongly negative to strongly positive) interact with dispersal distance to modulate the effects of fragmentation on species richness. Although this novel approach has already yielded some interesting results, some aspects might still need to be addressed.

### Positive demographic effects of fragmentation

An issue connected with the current approach is that by design, higher fragmentation landscapes have much higher dispersal mortality. Since the direction of dispersal is random, the probability of newly born individuals dispersing outside of the habitat is much higher in fragmented landscapes. This means that even if edge effects are strongly positive, they are outweighed by the negative effects of dispersal mortality, leading to strong negative demographic effects. 

For better understanding the spatial components of fragmentation effects, we are also interested in modelling positive demographic effects. In order to do so, we had the idea to implement a dispersal mode which forces dispersal to occur only within the habitat, which nullifies the risk of dispersal mortality. Such a dispersal mode might be less realistic for sessile organisms, although it can be a plausible scenario for plants with animal-mediated dispersal (e.g. seeds dispersed by birds), when the habitats of the dispersal vector and of the plant overlap.

The new dispersal mode is implemented in the `random_disperse` function in `Model/src/disperse.R`. The function draws a random cell in the landscape, and by setting `force_habitat = TRUE`, it will only draw from all habitat cells. That means that dispersal is not only restricted to habitat cells, but also detached from the original location of the parent. We will explore what implications this might have in the following analysis.

## Simulation setup

To explore the effects of random dispersal within habitat, we run a total of 9 simulations with the new dispersal mode (hereafter called "random_habitat") and 9 simulations with the same parameter combinations but the old dispersal mode ("short"). For each dispersal mode, every simulation has a unique combination of fragmentation (0.2, 0.5, 0.8), environmental autocorrelation (0, 0.5, 1). All other parameters are kept constant across simulations. The parameter combinations are summarized in the table below.

```{r}
#| label: tbl-params
#| tbl-cap: "Parameter combinations for the 18 simulations used in this analysis. The first 9 simulations use random dispersal within habitat, the last 9 simulations use short exponential dispersal with mean distance 2. 'fragmentation' refers to the spatial configuration of patches, with low values indicating fewer, bigger patches and high values indicating more, smaller patches; 'ac_amount' is the amount of environmental autocorrelation, 'habitat' is the proportion of habitat after fragmentation, and 'dispersal_dist' is the mean dispersal distance (only relevant for short dispersal)."
#| code-fold: true
library(here)
library(data.table)
library(dplyr)
log <- fread(here("output", "simulations_log.csv"))

params <- log[1:18, ] |> 
  select(sim_id, fragmentation, ac_amount, habitat, dispersal_type, dispersal_dist)

knitr::kable(params)
```


## Survey of population fluctuations

To explore the effects of dispersal mode on population fluctuations, we will first summarize the population sizes right before and after fragmentation and at the end of simulation for all 9x2 simulations. 

```{r}
#| label: tbl-populations
#| tbl-cap: "Population sizes after initialization ('start'), before fragmentation ('pre_fragmentation'), after fragmentation ('post_fragmentation') and at the end of simulation ('final') for all 18 simulations."
#| code-fold: true
library(purrr)
source("R/sim_select.R")

count_agents <- function(state) {
  steps <- names(state)
  populations <- sapply(steps, function(step) nrow(state[[step]]$agents))
  SR <- sapply(steps, function(step) length(unique(state[[step]]$agents$species_id)))
  return(data.frame(step = steps, population = populations, species_richness = SR))
}

r_paths <- sim_select(dispersal_type == "random", file_type = "state")
r_states <- purrr::map(here(r_paths), readRDS)
r_counts <- purrr::map(r_states, count_agents)

s_paths <- sim_select(dispersal_type == "short_long", file_type = "state")
s_states <- purrr::map(here(s_paths), readRDS)
s_counts <- purrr::map(s_states, count_agents)

params <- params |> 
  select(sim_id, dispersal_type, fragmentation, ac_amount)

# For each simulation, create df with steps as columns and populations as values
r_pops <- purrr::map2(r_counts, params$sim_id[params$dispersal_type == "random"], function(counts, sim_id) {
  counts$sim_id <- sim_id
  return(counts)
}) |> 
  bind_rows() |> 
  left_join(params, by = "sim_id") |> 
  select(-species_richness)

r_pops_wide <- tidyr::pivot_wider(names_from = step, values_from = population)

s_pops <- purrr::map2(s_counts, params$sim_id[params$dispersal_type == "short_long"], function(counts, sim_id) {
  counts$sim_id <- sim_id
  return(counts)
}) |> 
  bind_rows() |> 
  left_join(params, by = "sim_id") |> 
  select(-species_richness)

s_pops_wide <- tidyr::pivot_wider(names_from = step, values_from = population)

knitr::kable(rbind(r_pops_wide, s_pops_wide))
```

```{r}
#| label: fig-populations
#| fig-cap: "Population sizes after initialization ('start'), before fragmentation ('pre_fragmentation'), after fragmentation ('post_fragmentation') and at the end of simulation ('final') for all 18 simulations. The first 9 simulations use random dispersal within habitat, the last 9 simulations use short exponential dispersal with mean distance 2."
#| code-fold: true
library(ggplot2)
library(ggh4x)

pops <- rbind(r_pops, s_pops) |> 
  mutate(step = factor(step, levels = c("start", "pre_fragmentation", "post_fragmentation", "final")))

ggplot(pops, aes(x = step, y = population, group = sim_id, color = dispersal_type)) +
  geom_line() +
  facet_nested("Fragmentation" + fragmentation ~ "Autocorrelation" + ac_amount) +
  labs(x = "Simulation step", y = "Population size", color = "Dispersal mode") +
  theme_bw() +
  theme(legend.position = "bottom") 

```

### Dynamics of short dispersal

As we have seen, as long as niche breadth is narrow (0.1), short dispersal shows much higher recruitment and population sizes. Surprisingly, this is also the case before fragmentation for simulations on a landscape with no environmental autocorrelation (0), where we wouldn't expect significant differences. A possible explanation for this is that short dispersal might regularly lead to dispersal to the very same cell of the parent agent, which is virtually impossible for random dispersal (1/2500 before fragmentation). Since the origin cell is always suitable for the offspring, this would lead to higher recruitment and population sizes.
To test this, we will assess the probability of dispersal to the same cell for both dispersal modes. 

```{r}
source("Model/src/disperse.R")

s_count <- 0
r_count <- 0
n_sims <- 10000
for (i in seq(n_sims)) {
  cur_loc <- c(sample.int(50, 1), sample.int(50, 1))

  new_loc_s <- toroidal_disperse(
    cur_loc = cur_loc,
    d_sd = 2,
    d_mean = 2,
    grid_size = 50,
    kernel = "exponential"
  )
  s_count <- s_count + all(new_loc_s == cur_loc)

  new_loc_r <- random_disperse()
  r_count <- r_count + all(new_loc_r == cur_loc)
}
cat(
  "Probability of dispersing to the same cell:\n",
  "short dispersal (mean distance = 2):",
  s_count / n_sims * 100,
  "%\n",
  "random dispersal:",
  r_count / n_sims * 100,
  "%\n"
)
```

As expected, almost 1 in 4 dispersal events with short dispersal lead to dispersal to the same cell, while this is virtually impossible for random dispersal. This means that even without environmental autocorrelation, short dispersal is highly advantageous for recruitment and population sizes, especially when niche breadth is narrow. But how does this change on the fragmented landscape? Let's first look at the growth rates of the populations after fragmentation.



To better highlight the differences, we can also investigate the actual probabilities of recruitment on a fragmented landscape.

We start by initializing a full habitat landscape with medium autocorrelation, populated with 10000 agents from 1000 species.
```{r}
library(data.table)
library(raster)
source("Model/src/initialize.R")
source("Model/src/generate_agents.R")
source("Model/src/distribute_agents.R")
source("Model/src/landscape.R")
source("Model/src/record_state.R")

species_par <- data.frame(
  species_id = 1:1000,
  n_value = seq(from = 0, to = 1, length.out = 1000),
  birth_rate = 0.85,
  death_rate = 0.25
)

state_init <- initialize(
  grid_size = 50,
  ac_amount = 0.5,
  n_species = 1000,
  n_pop = 10000,
  species_par = species_par,
  k_inter = 50,
  k_intra = 50,
  niche_breadth = 0.1,
  state_type = "full_state",
  master_seed = 123
)
```
Note that `distribute_agents` draws a suitable cell first and checks if cell capacity is reached only after, which means that the actual number of agents distributed might be lower than the requested number (especially for big populations). In this case `{r} nrow(state_init$agents)` agents were distributed. 

We can now apply different levels of fragmentation to the landscape and record the state of the population after fragmentation.
```{r}
source("Model/src/fragmentation.R")
frag_low <- fragment(full_state = state_init, habitat = 0.15, fragmentation = 0.2)
frag_medium <- fragment(full_state = state_init, habitat = 0.15, fragmentation = 0.5)
frag_high <- fragment(full_state = state_init, habitat = 0.15, fragmentation = 0.8)
```

Now, we can pick a random agent and assess the probability of recruitment for both dispersal modes. We will do this by simulating 1000 birth events for randomly picked agents and counting how many of them lead to successful recruitment.

```{r}
#| code-fold: true

source("Model/src/birth.R")

n_sims <- 10000
success_short <- list(low = 0, medium = 0, high = 0)
success_random <- list(low = 0, medium = 0, high = 0)

nb <- 0.1

for (i in seq(n_sims)) {

  # Low frag
  random_agent_low <- frag_low$agents[sample.int(nrow(frag_low$agents), 1), ]
  u <- species_par$n_value[random_agent_low$species_id]
  cur_loc_low <- c(random_agent_low$x_loc, random_agent_low$y_loc)
  new_loc_s_low <- toroidal_disperse(
    cur_loc = cur_loc_low,
    d_sd = 2,
    d_mean = 2,
    grid_size = 50,
    kernel = "exponential"
  )
  new_loc_r_low <- random_disperse(
    nrow = 50,
    ncol = 50,
    habitat_cells = frag_low$cells$habitat,
    force_habitat = TRUE
  )
  if (check_recruit(agents = frag_low$agents, 
                    grid = frag_low$grid, 
                    location = new_loc_s_low, 
                    u = u, 
                    nb = nb,
                    species_id = random_agent_low$species_id,
                    k_inter = 50,
                    k_intra = 50)) {
    success_short$low <- success_short$low + 1
  }
  if (check_recruit(agents = frag_low$agents, 
                    grid = frag_low$grid, 
                    location = new_loc_r_low, 
                    u = u,
                    nb = nb,
                    species_id = random_agent_low$species_id,
                    k_inter = 50,
                    k_intra = 50)) {
    success_random$low <- success_random$low + 1
  }

  # Medium frag
  random_agent_medium <- frag_medium$agents[sample.int(nrow(frag_medium$agents), 1), ]
  u <- species_par$n_value[random_agent_medium$species_id]
  cur_loc_medium <- c(random_agent_medium$x_loc, random_agent_medium$y_loc)
  new_loc_s_medium <- toroidal_disperse(
    cur_loc = cur_loc_medium,
    d_sd = 2,
    d_mean = 2,
    grid_size = 50,
    kernel = "exponential"
  )
  new_loc_r_medium <- random_disperse(
    nrow = 50,
    ncol = 50,
    habitat_cells = frag_medium$cells$habitat,
    force_habitat = TRUE
  )
  if (check_recruit(agents = frag_medium$agents, 
                    grid = frag_medium$grid, 
                    location = new_loc_s_medium, 
                    u = u,
                    nb = nb,
                    species_id = random_agent_medium$species_id,
                    k_inter = 50,
                    k_intra = 50)) {
    success_short$medium <- success_short$medium + 1
  }
  if (check_recruit(agents = frag_medium$agents, 
                    grid = frag_medium$grid, 
                    location = new_loc_r_medium, 
                    u = u,
                    nb = nb,
                    species_id = random_agent_medium$species_id,
                    k_inter = 50,
                    k_intra = 50)) {
    success_random$medium <- success_random$medium + 1
  }

  # High frag
  random_agent_high <- frag_high$agents[sample.int(nrow(frag_high$agents), 1), ]
  u <- species_par$n_value[random_agent_high$species_id]
  cur_loc_high <- c(random_agent_high$x_loc, random_agent_high$y_loc)
  new_loc_s_high <- toroidal_disperse(
    cur_loc = cur_loc_high,
    d_sd = 2,
    d_mean = 2,
    grid_size = 50,
    kernel = "exponential"
  )
  new_loc_r_high <- random_disperse(
    nrow = 50,
    ncol = 50,
    habitat_cells = frag_high$cells$habitat,
    force_habitat = TRUE
  )
  if (check_recruit(agents = frag_high$agents, 
                    grid = frag_high$grid, 
                    location = new_loc_s_high, 
                    u = u,
                    nb = nb,
                    species_id = random_agent_high$species_id,
                    k_inter = 50,
                    k_intra = 50)) {
    success_short$high <- success_short$high + 1
  }
  if (check_recruit(agents = frag_high$agents, 
                    grid = frag_high$grid, 
                    location = new_loc_r_high, 
                    u = u,
                    nb = nb,
                    species_id = random_agent_high$species_id,
                    k_inter = 50,
                    k_intra = 50)) {
    success_random$high <- success_random$high + 1
  }
}

cat(
  "Recruitment success probabilities with nb =", nb, "\n\n",
  "Short dispersal (mean distance = 2):\n",
  "   Low fragmentation:", success_short$low / n_sims * 100, "%\n",
  "   Medium fragmentation:", success_short$medium / n_sims * 100, "%\n",
  "   High fragmentation:", success_short$high / n_sims * 100, "%\n\n",
  "Random dispersal within habitat:\n",
  "   Low fragmentation:", success_random$low / n_sims * 100, "%\n",
  "   Medium fragmentation:", success_random$medium / n_sims * 100, "%\n",
  "   High fragmentation:", success_random$high / n_sims * 100, "%\n"
)

```