---
title: "Distance Decay Analysis"
author: "J. N. Thiem"
bibliography: fragmentation.bib
csl: apa-eu.csl
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    respect-user-color-scheme: true
    theme: 
      light: flatly
      dark: darkly
    highlight-style: 
      light: github-light
      dark: github-dark
execute:
  echo: true
  warning: false
  message: false
---

## Introduction {#sec-intro}
In this analysis, I will explore applications of distance decay curves to different simulated datasets. Distance decay analyses are used to understand how similarity between communities changes with geographic distance. 
The driving idea is that distance decay curves can be a compelling tool to discern between geometric and demographic effects of fragmentation. If we isolate the geometric component of fragmentation effects, that is if we sample data right after fragmentation, before any ecological processes have taken place on the fragmented landscape, the relationship between community similarity and geographic distance should be consistent, regardless of the spatial configuration of fragmentation.
This should not be the case if we include demographic effects into the equation, when we sample our data after some time has passed since fragmentation, and ecological processes have had the chance to play out. In this case, we should expect to see a stronger distance decay of similarity in the more fragmented landscape, where spatial isolation of communities is more pronounced (?).

All data used for this demonstration was generated using the same model version (1.1) and sampled using the checkerboard method.

Following parameters were kept constant across all simulations:

* Landscape size: 50 x 50 grid cells
* Number of time steps: 100 
    + 40 steps for the burn-in phase
    + 60 steps after fragmentation
* Initial number of species: 1000
* Initial number of individuals: 10000
* Dispersal kernel: negative exponential

## Fragmentation as the only driver {#sec-frag-only}
To test the method, I started by running a relative large number of simulations where all ecological parameters were kept constant, except for the spatial configuration of fragmentation which was varied in 3 different levels: 0.2 (low), 0.5 (medium), 0.8 (high).

All simulations for this section were performed with the following parameters:

* Landscape autocorrelation: 0.7
* Habitat amount (after fragmentation): 0.15
* Edge effect: none
* Mean dispersal distance: 2 (in units of grid cells)

```{r}
#| label: fig-1
#| fig-cap: "Fragmentation effects on distance decay and species richness. A: Distance decay curves for different levels of fragmentation, sampled right after fragmentation (left) and at the end of the simulation (right). Shaded areas represent 95% confidence intervals across simulations. B: Species richness for different levels of fragmentation, sampled right after fragmentation (left column) and at the end of the simulation (right column). Richness was either averaged across all samples from each simulation (top row) or assessed across the whole landscape (bottom row)."
#| fig-width: 10
#| fig-height: 10
library(here)
library(data.table)
library(tidyverse)
library(patchwork)
source(here("R", "sim_select.R"))
source(here("R", "dist_decay.R"))

# All simulations with only fragmentation as variable parameter, checkerboard sampling design
paths <- sim_select(sim_id <= 75, sampled = "checkerboard")
data_list <- purrr::map(here(paths), fread)

# Convert to wide format for distance decay analysis
data_wide <- data_list %>%
    map(
        ~ .x %>%
            tidyr::pivot_wider(
            names_from = species_id,
            values_from = n,
            values_fill = 0,
            names_prefix = "sp_",
            names_sort = TRUE
    )
)

# Run distance decay analysis for each simulation
dd_results <- list(
    post_frag = vector("list", length(data_wide)),
    final = vector("list", length(data_wide))
)
for (i in seq_along(data_wide)) {
    
    sim_id <- data_wide[[i]]$sim_id[1]
    
    post_frag <- data_wide[[i]] %>%
        dplyr::filter(step_label == "post_fragmentation")

    final <- data_wide[[i]] %>%
    dplyr::filter(step_label == "final")

    fragmentation <- final$fragmentation[1]

    dd_results$post_frag[[i]] <- dist_decay(
        post_frag,
        dist_type = "euclidean",
        distvec = seq(0, 25, length.out = 200)
    )$smooth %>%
        dplyr::select(distance, similarity) %>%
        dplyr::mutate(
            sim_id = sim_id, 
            fragmentation = fragmentation, 
            step_label = "post_fragmentation"
        )

    dd_results$final[[i]] <- dist_decay(
        final,
        dist_type = "euclidean",
        distvec = seq(0, 25, length.out = 200)
    )$smooth %>%
        dplyr::select(distance, similarity) %>%
        dplyr::mutate(
            sim_id = sim_id, 
            fragmentation = fragmentation, 
            step_label = "final"
        )
}

# Average distance decay curves across time step for each fragmentation level
dd_summaries <- map(dd_results, ~ .x %>%
    bind_rows() %>%
    group_by(step_label, fragmentation, distance) %>%
    summarise(
        simi_low = quantile(similarity, 0.025, na.rm = TRUE),
        simi_high = quantile(similarity, 0.975, na.rm = TRUE),
        similarity = mean(similarity, na.rm = TRUE), .groups = "drop"
    ) %>%
    mutate(
        fragmentation = factor(
            fragmentation, 
            levels = c(0.2, 0.5, 0.8),
            labels = c("Low", "Medium", "High")
        )
    )
)

dd_summary <- bind_rows(dd_summaries) %>% 
    mutate(
        step_label = factor(
            step_label, 
            levels = c("post_fragmentation", "final"), 
            labels = c("Post-fragmentation", "End of simulation")
        )
    )

# Assess alpha and gamma div for each dataset
SR_results <- list(
    post_frag = vector("list", length(data_list)),
    final = vector("list", length(data_list))
)
for (i in seq_along(data_list)) {

    post_frag <- data_list[[i]] %>% 
        dplyr::filter(step_label == "post_fragmentation")

    pf_summary <- post_frag %>% 
        dplyr::group_by(sim_id, fragmentation, sample_id) %>% 
        dplyr::summarise(richness = n(), .groups = "drop")
    
    pf_alpha <- mean(pf_summary$richness, na.rm = TRUE)
    pf_gamma <- length(unique(post_frag$species_id))
    
    SR_results$post_frag[[i]] <- list(summary = pf_summary,
                                      alpha = pf_alpha,
                                      gamma = pf_gamma)
    
    final <- data_list[[i]] %>% 
        dplyr::filter(step_label == "final")
    
    fin_summary <- final %>% 
        dplyr::group_by(sim_id, fragmentation, sample_id) %>% 
        dplyr::summarise(richness = n(), .groups = "drop")

    fin_alpha <- mean(fin_summary$richness, na.rm = TRUE)
    fin_gamma <- length(unique(final$species_id))

    sim_id <- first(final$sim_id)
    fragmentation <- first(final$fragmentation)

    SR_results$final[[i]] <- list(sim_id = sim_id,
                                  fragmentation = fragmentation,
                                  summary = fin_summary,
                                  alpha = fin_alpha,
                                  gamma = fin_gamma)

}

# Aggregate species richness data into tidy format for plotting
# Alpha diversity (local scale) - per sample
alpha_tidy <- bind_rows(
    # Post-fragmentation
    SR_results$post_frag %>%
        map_df(
        ~ .x$summary %>%
            mutate(step_label = "post_fragmentation"),
        .id = "dataset_idx"
        ),
    
    # Final
    SR_results$final %>%
        map_df(
        ~ .x$summary %>%
            mutate(step_label = "final"),
        .id = "dataset_idx"
        )
) %>%
    mutate(scale = "Local (alpha)",
           richness_value = richness) %>%
    select(sim_id, fragmentation, step_label, scale, richness_value)

# Gamma diversity (landscape scale) - per dataset
gamma_tidy <- bind_rows(
    # Post-fragmentation
    SR_results$post_frag %>%
        map_df(
            ~ {
                tibble(
                    sim_id = .x$summary$sim_id[1],
                    fragmentation = .x$summary$fragmentation[1],
                    step_label = "post_fragmentation",
                    scale = "Landscape (gamma)",
                    richness_value = .x$gamma
                )
            }
        ),
  
    # Final
    SR_results$final %>%
        map_df(
            ~ {
                tibble(
                    sim_id = .x$sim_id,
                    fragmentation = .x$summary$fragmentation[1],
                    step_label = "final",
                    scale = "Landscape (gamma)",
                    richness_value = .x$gamma
                )
            }
        )
)

# Combine alpha and gamma
sr_combined <- bind_rows(alpha_tidy, gamma_tidy) %>%
    mutate(
        fragmentation = factor(
            fragmentation, 
            levels = c(0.2, 0.5, 0.8), 
            labels = c("Low", "Medium", "High")
        ),
        step_label = factor(
            step_label, 
            levels = c("post_fragmentation", "final"), 
            labels = c("Post-fragmentation", "End of simulation")
        ),
        scale = factor(scale, levels = c("Local (alpha)", "Landscape (gamma)")))

# Plot distance decay curves
pal <- c("midnightblue", "darkgoldenrod", "violetred4")
gg_dd <- ggplot(dd_summary, aes(x = distance, y = similarity, color = fragmentation, fill = fragmentation)) +
    geom_line(linewidth = 1, alpha = 0.8) +
    geom_ribbon(aes(ymin = simi_low, ymax = simi_high), alpha = 0.2, color = NA) +
    facet_wrap(~ step_label) +
    scale_color_manual(values = pal) +
    scale_fill_manual(values = pal) +
    labs(x = "Euclidean Distance", y = "Similarity (1 - Bray-Dissimilarity)", tag = "A") +
    theme_bw(base_size = 14) +
    theme(legend.position = "top")

# Plot species richness with violin plots
gg_sr <- ggplot(sr_combined, aes(x = fragmentation, y = richness_value, color = fragmentation, fill = fragmentation)) +
    geom_violin(alpha = 0.5) +
    geom_jitter(width = 0.1, alpha = 0.3, size = 1) +
    facet_grid(scale ~ step_label, scales = "free_y") +
    scale_color_manual(values = pal) +
    scale_fill_manual(values = pal) +
    labs(x = "Fragmentation Level", y = "Species Richness", tag = "B") +
    theme_bw(base_size = 14) +
    theme(legend.position = "none", strip.text.x = element_blank())

gg_dd / gg_sr

```

As we can see, although distance decay was averaged for a total of 22 simulations per fragmentation level, the curves overlap almost perfectly when we isolate geometric effects only (right after fragmentation). In contrast, when looking at the net effects at the end of the simulation, we can see clear differences between the different levels of fragmentation. Higher fragmentation scenarios (i.e. many small habitat patches) show higher community similarity at low distances but a steeper decrease in similarity at increasing distance than low fragmentation scenarios (i.e. fewer yet bigger habitat patches). 

The plot also shows the differences between fragmentation levels in terms of species richness. In this case, we can see that geometric effects of fragmentation are neutral at the local scale but positive at the landscape scale, while net effects are negative at the local scale and neutral at the landscape scale. This aligns with what was shown and discussed by Gelber et al. [@gelber_geometric_2025].

## Additional drivers {#sec-add-drivers}

So far so good, but what happens if we look at different scenarios where also other ecological parameters are changed? 

As described before, the simulations I analysed so far only differ in fragmentation configuration while all ecological parameters were fixed across all simulations. This approach was useful to highlight the direct effects of fragmentation itself, but it has the downside that it only describes a very specific scenario, with relatively high landscape autocorrelation, a small amount of habitat left after fragmentation, no edge effects and short dispersal of species (see @sec-frag-only).

In the following analyses, I will use simulations where one additional parameter at a time has been varied (alongside fragmentation). Alongside the 3 levels of fragmentation already used above (0.2, 0.5, 0.8), each parameter has been varied into 3 different levels as well (the exact levels used will be mentioned in the respective subsections). I will run the same type of analysis shown in the previous section, just using data sampled from different simulations.

Note that so far I only run and sampled a single simulation for each of the following parameter combination, for a total of 9 simulations per additional driver. The following sections are solely for demonstration purposes. For a thorough analysis, many more replicates for each scenario will be needed.


### Habitat amount {#sec-habitat}

The amount of habitat preserved after the fragmentation process can affect quite strongly how different configurations shape diversity and composition of communities. Especially when dispersal is limited to short distances, as it is the case in my simulations, the negative net effects of higher fragmentation on alpha diversity (as in local species richness) as well as the steeper decline of similarity at increasing distances should be partly buffered if a larger amount of habitat is preserved.

To investigate the effects of habitat amount on the relationship between fragmentation and diversity, 3 levels of habitat amount were simulated (0.1, 0.3, 0.5). 

```{r}
#| label: fig-2
#| fig-cap: "Fragmentation effects on distance decay of similarity for different habitat amounts. Distance decay curves have different colors for different levels of fragmentation, sampled right after fragmentation (left column) and at the end of the simulation (right column). The amount of habitat preserved after fragmentation is shown increasing from top to bottom. Shaded areas represent 95% confidence intervals of the distance decay curves." 
#| fig-width: 10
#| fig-height: 10

# All simulations with habitat %in% c(0.1, 0.3, 0.5)
paths <- sim_select(
    habitat %in% c(0.1, 0.3, 0.5), 
    sampled = "checkerboard"
)
data_list <- purrr::map(here(paths), fread)

# Convert to wide format for distance decay analysis
data_wide <- data_list %>%
    map(
        ~ .x %>%
            tidyr::pivot_wider(
            names_from = species_id,
            values_from = n,
            values_fill = 0,
            names_prefix = "sp_",
            names_sort = TRUE
    )
)

# Run distance decay analysis for each simulation
dd_results <- list(
    post_frag = vector("list", length(data_wide)),
    final = vector("list", length(data_wide))
)
for (i in seq_along(data_wide)) {
    
    sim_id <- data_wide[[i]]$sim_id[1]
    
    post_frag <- data_wide[[i]] %>%
        dplyr::filter(step_label == "post_fragmentation")

    final <- data_wide[[i]] %>%
    dplyr::filter(step_label == "final")

    fragmentation <- first(final$fragmentation)
    habitat <- first(final$habitat)

    dd_results$post_frag[[i]] <- dist_decay(
        post_frag,
        dist_type = "euclidean",
        distvec = seq(0, 25, length.out = 200)
    )$smooth %>%
        dplyr::mutate(
            sim_id = sim_id, 
            fragmentation = factor(
                fragmentation, 
                levels = c(0.2, 0.5, 0.8),
                labels = c("Low", "Medium", "High")
            ),
            habitat = factor(
                habitat,
                levels = c(0.1, 0.3, 0.5),
                labels = c("10% habitat", "30% habitat", "50% habitat")
            ),
            step_label = "post_fragmentation"
        )

    dd_results$final[[i]] <- dist_decay(
        final,
        dist_type = "euclidean",
        distvec = seq(0, 25, length.out = 200)
    )$smooth %>%
        dplyr::mutate(
            sim_id = sim_id, 
            fragmentation = factor(
                fragmentation, 
                levels = c(0.2, 0.5, 0.8),
                labels = c("Low", "Medium", "High")
            ),            
            habitat = factor(
                habitat,
                levels = c(0.1, 0.3, 0.5),
                labels = c("10% habitat", "30% habitat", "50% habitat")
            ),
            step_label = "final"
        )
}

dd_summary <- bind_rows(dd_results) %>% 
    mutate(
        step_label = factor(
            step_label, 
            levels = c("post_fragmentation", "final"), 
            labels = c("Post-fragmentation", "End of simulation")
        )
    )

ggplot(dd_summary, aes(x = distance, y = similarity, color = fragmentation, fill = fragmentation)) +
    geom_line(linewidth = 1, alpha = 0.8) +
    geom_ribbon(aes(ymin = simi_low, ymax = simi_high), alpha = 0.2, color = NA) +
    facet_grid(rows = vars(habitat), cols = vars(step_label)) +
    scale_color_manual(values = pal) +
    scale_fill_manual(values = pal) +
    labs(x = "Euclidean Distance", y = "Similarity (1 - Bray-Dissimilarity)") +
    theme_bw(base_size = 14) +
    theme(legend.position = "bottom")
```

### Edge effects {#sec-edge}

```{r}
#| label: fig-3
#| fig-cap: "Fragmentation effects on distance decay of similarity for different edge effects. Distance decay curves have different colors for different levels of fragmentation. Samples were taken right after fragmentation (left column) and at the end of the simulation (right column). Edge effects are either negative (top row), neutral (middle row) or positive (bottom row). Shaded areas represent 95% confidence intervals of the distance decay curves."
#| fig-width: 10
#| fig-height: 10

paths <- sim_select(
    sim_id > 84 & sim_id <= 93, 
    sampled = "checkerboard"
)

data_list <- purrr::map(here(paths), fread)

data_wide <- data_list %>%
    map(
        ~ .x %>%
            tidyr::pivot_wider(
            names_from = species_id,
            values_from = n,
            values_fill = 0,
            names_prefix = "sp_",
            names_sort = TRUE
    )
)

dd_results <- list(
    post_frag = vector("list", length(data_wide)),
    final = vector("list", length(data_wide))
)
for (i in seq_along(data_wide)) {
    
    sim_id <- data_wide[[i]]$sim_id[1]
    
    post_frag <- data_wide[[i]] %>%
        dplyr::filter(step_label == "post_fragmentation")

    final <- data_wide[[i]] %>%
    dplyr::filter(step_label == "final")

    fragmentation <- first(final$fragmentation)
    edge_effect <- first(final$edge_effect)

    dd_results$post_frag[[i]] <- dist_decay(
        post_frag,
        dist_type = "euclidean",
        distvec = seq(0, 25, length.out = 200)
    )$smooth %>%
        dplyr::mutate(
            sim_id = sim_id, 
            fragmentation = factor(
                fragmentation, 
                levels = c(0.2, 0.5, 0.8),
                labels = c("Low", "Medium", "High")
            ),
            edge_effect = factor(
                edge_effect,
                levels = c(0.5, 1, 2),
                labels = c("Negative", "None", "Positive")
            ),
            step_label = "post_fragmentation"
        )

    dd_results$final[[i]] <- dist_decay(
        final,
        dist_type = "euclidean",
        distvec = seq(0, 25, length.out = 200)
    )$smooth %>%
        dplyr::mutate(
            sim_id = sim_id, 
            fragmentation = factor(
                fragmentation, 
                levels = c(0.2, 0.5, 0.8),
                labels = c("Low", "Medium", "High")
            ),            
            edge_effect = factor(
                edge_effect,
                levels = c(0.5, 1, 2),
                labels = c("Negative", "None", "Positive")
            ),
            step_label = "final"
        )
}

dd_summary <- bind_rows(dd_results) %>% 
    mutate(
        step_label = factor(
            step_label, 
            levels = c("post_fragmentation", "final"), 
            labels = c("Post-fragmentation", "End of simulation")
        )
    )

gg_edge <- ggplot(dd_summary, aes(x = distance, y = similarity, color = fragmentation, fill = fragmentation)) +
    geom_line(linewidth = 1, alpha = 0.8) +
    geom_ribbon(aes(ymin = simi_low, ymax = simi_high), alpha = 0.2, color = NA) +
    facet_grid(rows = vars(edge_effect), cols = vars(step_label)) +
    scale_color_manual(values = pal) +
    scale_fill_manual(values = pal) +
    labs(x = "Euclidean Distance", y = "Similarity (1 - Bray-Dissimilarity)") +
    theme_bw(base_size = 14) +
    theme(legend.position = "bottom")

gridExtra::grid.arrange(gg_edge, right = "Edge effects")

```


### Dispersal distance {#sec-dispersal}

The mean dispersal distance of species can also affect how fragmentation shapes diversity and composition of communities. Similarly to habitat amount, when dispersal is limited to short distances, the negative net effects of higher fragmentation on alpha diversity (as in local species richness) as well as the steeper decline of similarity at increasing distances should be more pronounced than when dispersal is less limited. To investigate the effects of dispersal distance on the relationship between fragmentation and diversity, 3 levels of mean dispersal distance were used for different scenarios (1, 5, 10 grid cells).

```{r}
#| label: fig-4
#| eval: false
#| fig-cap: "Fragmentation effects on distance decay of similarity for different mean dispersal distances. Distance decay curves have different colors for different levels of fragmentation. Samples were taken right after fragmentation (left column) and at the end of the simulation (right column). The mean dispersal distance of species is shown increasing from top to bottom. Shaded areas represent 95% confidence intervals of the distance decay curves."
#| fig-width: 10
#| fig-height: 10

paths <- sim_select(
    dispersal_dist %in% c(1, 5, 10), 
    sampled = "checkerboard"
)
data_list <- purrr::map(here(paths), fread)

data_wide <- data_list %>%
    map(
        ~ .x %>%
            tidyr::pivot_wider(
            names_from = species_id,
            values_from = n,
            values_fill = 0,
            names_prefix = "sp_",
            names_sort = TRUE
    )
)

dd_results <- list(
    post_frag = vector("list", length(data_wide)),
    final = vector("list", length(data_wide))
)
for (i in seq_along(data_wide)) {
    
    sim_id <- data_wide[[i]]$sim_id[1]
    
    post_frag <- data_wide[[i]] %>%
        dplyr::filter(step_label == "post_fragmentation")

    final <- data_wide[[i]] %>%
    dplyr::filter(step_label == "final")

    fragmentation <- first(final$fragmentation)
    dispersal_dist <- first(final$dispersal_dist)

    dd_results$post_frag[[i]] <- dist_decay(
        post_frag,
        dist_type = "euclidean",
        distvec = seq(0, 25, length.out = 200)
    )$smooth %>%
        dplyr::mutate(
            sim_id = sim_id, 
            fragmentation = factor(
                fragmentation, 
                levels = c(0.2, 0.5, 0.8),
                labels = c("Low", "Medium", "High")
            ),
            dispersal_dist = factor(dispersal_dist),
            step_label = "post_fragmentation"
        )

    dd_results$final[[i]] <- dist_decay(
        final,
        dist_type = "euclidean",
        distvec = seq(0, 25, length.out = 200)
    )$smooth %>%
        dplyr::mutate(
            sim_id = sim_id, 
            fragmentation = factor(
                fragmentation, 
                levels = c(0.2, 0.5, 0.8),
                labels = c("Low", "Medium", "High")
            ),            
            dispersal_dist = factor(dispersal_dist),
            step_label = "final"
        )
}

dd_summary <- bind_rows(dd_results) %>% 
    mutate(
        step_label = factor(
            step_label, 
            levels = c("post_fragmentation", "final"), 
            labels = c("Post-fragmentation", "End of simulation")
        )
    )

gg_disp <- ggplot(dd_summary, aes(x = distance, y = similarity, color = fragmentation, fill = fragmentation)) +
    geom_line(linewidth = 1, alpha = 0.8) +
    geom_ribbon(aes(ymin = simi_low, ymax = simi_high), alpha = 0.2, color = NA) +
    facet_grid(rows = vars(dispersal_dist), cols = vars(step_label)) +
    scale_color_manual(values = pal) +
    scale_fill_manual(values = pal) +
    labs(x = "Euclidean Distance", y = "Similarity (1 - Bray-Dissimilarity)") +
    theme_bw(base_size = 14) +
    theme(legend.position = "bottom") 
    
gridExtra::grid.arrange(gg_disp, right = "Mean dispersal distance")

```





